---
title: "Global dynamics of population, land use, and inequality"
author: "Scott Claessens"
format: pdf
---

# Overview

...

# Datasets

...

# Causal model

Static DAG:

```{r}
library(dagitty)

dagitty('dag{
  P [pos = "0,0"]
  C [pos = "1,-1"]
  G [pos = "2,0"]
  G <- P ;
  G <- C ;
  C <- P ;
}') |> plot()
```

Dynamic DAG:

```{r}
dagitty('dag{
  P0 [pos = "0,0"]
  P1 [pos = "1,0"]
  C0 [pos = "0,-1"]
  C1 [pos = "1,-1"]
  G0 [pos = "0,-2"]
  G1 [pos = "1,-2"]
  P1 <- P0 ;
  C1 <- C0 ;
  G1 <- G0 ;
  C1 <- P0 ;
  G1 <- P0 ;
  G1 <- C0 ;
}') |> plot()
```

# Statistical model

Likelihoods:

$$\begin{align}
P &\sim \text{HurdleLogNormal}(p_P, \mu_P, \sigma) \\
C &\sim \text{HurdleLogNormal}(p_C, \mu_C, \omega) \\
G &\sim \text{BetaProportion}(\mu_G, \phi)
\end{align}$$

Differential equations:

$$\frac{d\text{logit}(p_P)}{dt} = e^{b_P} \\
\frac{d\mu_P}{dt} = e^{r_P} \\
\frac{d\text{logit}(p_C)}{dt} = e^{b_C} \\
\frac{d\mu_C}{dt} = e^{r_P}\frac{P}{C} \\
\frac{d\text{logit}(\mu_G)}{dt} = r_G$$

where:

$$\begin{align}
r_G &= \alpha \\
&+ \beta_P p_P\text{log}(e^{\mu_P} + 1) \\
&+ \beta_C p_C\text{log}(e^{\mu_C} + 1) \\
&- e^{\gamma} \text{logit}(\mu_G)
\end{align}$$

Initial values:

$$\begin{align}
\text{logit}(p_P)(0) &\sim \mathcal{N}(-4, 0.5) \\
\mu_P(0) &\sim \mathcal{N}(-1, 0.5) \\
\text{logit}(p_C)(0) &\sim \mathcal{N}(-4, 0.5) \\
\mu_C(0)  &\sim \mathcal{N}(-1, 0.5) \\
\text{logit}(\mu_G)(0) &\sim \mathcal{N}(-1, 0.5)
\end{align}$$

In addition:

- Varying effects across subregions, with spatial GPs for $\alpha$ and $\text{logit}(\mu_G)(0)$. Varying effects have variance $\tau$ and GPs have length scale $lscale$
- Differential equations for $P$ and $C$ estimated piecewise across three time periods: (1) 10,000 BCE to 0 CE, (2) 0 CE to 1600 CE, and (3) 1600 CE to present

Priors:

$$\begin{align}
b_P &\sim \text{Normal}(-2, 0.5) \\
r_P &\sim \text{Normal}(-2, 0.5) \\
b_C &\sim \text{Normal}(-2, 0.5) \\
r_C &\sim \text{Normal}(-2, 0.5) \\
\alpha &\sim \text{Normal}(0, 0.5) \\
\beta_P &\sim \text{Normal}(0, 0.5) \\
\beta_C &\sim \text{Normal}(0, 0.5) \\
\gamma &\sim \text{Normal}(-2, 0.5) \\
\tau &\sim \text{Exponential}(2) \\
lscale &\sim \text{Exponential}(2) \\
\sigma &\sim \text{Exponential}(2) \\
\omega &\sim \text{Exponential}(2) \\
\phi &\sim \text{Exponential}(2) \\
\end{align} $$

# Results
